Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\tnoteref{t1,t2}

### 2016-03-10 curvature.tex
removed uneccesary page break
4 leaf treespace figure

rSPR operations give rise to a distance measure between $X$-trees: $\dspr{T_1, T_2}$ is the minimum number of rSPR operations required to transform an $X$-tree $T_1$ into $T_2$.
For example, the trees in Figure~\ref{fig:two-spr} are separated by two rSPR operations.
Moreover, rSPR operations naturally give rise to a graph on the set of $X$-trees for which this distance is simply the shortest-path graph distance.
Let $\mathcal{T}_n$ be the set of trees with $n$ leaves and label set $X = \set{1, 2, \ldots n, \rho}$.
Then the rSPR graph $G$ of $\mathcal{T}_n$ is the graph with vertex set $V(G) = \mathcal{T}_n$ and edge set $E(G) = \set{(T,S) \mid \dspr{T,S} = 1, T \in V, S \in V}$.

@@

The \emph{coarse Ricci-Ollivier curvature} of $x$ and $y$ is then defined as:
\vspace{-0.5em}
\begin{equation}
\curvature{m; x, y} := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}.
\label{eq:curvatureDef}
\end{equation}
For the purposes of this paper, ``curvature'' without further specification will refer to \eqref{eq:curvatureDef}.
We will use $\curvature{x,y}$ to denote the curvature of the simple (uniform choice of neighbor) random walk, and use $\curvature{\MH; x,y}$ to indicate curvature with respect to the Metropolis-Hastings random walk sampling the uniform distribution (described in detail in Section~\ref{sec:random_walks}).
Positive curvature implies that the neighborhoods $m_x$ and $m_y$ are closer in transportation distance than point masses at $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are more distant than point masses at $x$ and $y$.
Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with a random walk.
\begin{figure}[!h]
	\hspace*{\stretch{1}}
	\includegraphics[width=0.43\textwidth]{figs/neighborhood}
	\hspace*{\stretch{1}}
	\caption{The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections with $T$ and between neighbors.}
	\label{fig:neighborhood}
\end{figure}
\vspace{0.1em}

\pagebreak

>>>>

rSPR operations give rise to a distance measure between $X$-trees: $\dspr{T_1, T_2}$ is the minimum number of rSPR operations required to transform an $X$-tree $T_1$ into $T_2$.
For example, the trees in Figure~\ref{fig:two-spr} are separated by two rSPR operations.
Moreover, rSPR operations naturally give rise to a graph on the set of $X$-trees for which this distance is simply the shortest-path graph distance.
Let $\mathcal{T}_n$ be the set of trees with $n$ leaves and label set $X = \set{1, 2, \ldots n, \rho}$.
Then the rSPR graph $G$ of $\mathcal{T}_n$ is the graph with vertex set $V(G) = \mathcal{T}_n$ and edge set $E(G) = \set{(T,S) \mid \dspr{T,S} = 1, T \in V, S \in V}$.
Figure~\ref{fig:4-taxa-treespace} illustrates the rSPR graph of 4-leaf trees.

@@

The \emph{coarse Ricci-Ollivier curvature} of $x$ and $y$ is then defined as:
\vspace{-0.5em}
\begin{equation}
\curvature{m; x, y} := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}.
\label{eq:curvatureDef}
\end{equation}
For the purposes of this paper, ``curvature'' without further specification will refer to \eqref{eq:curvatureDef}.
We will use $\curvature{x,y}$ to denote the curvature of the simple (uniform choice of neighbor) random walk, and use $\curvature{\MH; x,y}$ to indicate curvature with respect to the Metropolis-Hastings random walk sampling the uniform distribution (described in detail in Section~\ref{sec:random_walks}).
Positive curvature implies that the neighborhoods $m_x$ and $m_y$ are closer in transportation distance than point masses at $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are more distant than point masses at $x$ and $y$.
Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with a random walk.
\begin{figure}
	\hspace*{\stretch{1}}
	\subfigure[\label{fig:4-taxa-treespace}]{\includegraphics[width=0.55\textwidth]{figs/4-taxa-treespace}}
	\hspace*{\stretch{2}}
	\subfigure[\label{fig:neighborhood}]{\includegraphics[width=0.4\textwidth]{figs/neighborhood}}
	\hspace*{\stretch{1}}
	\caption{(a) The rSPR graph of $X$-trees with 4 leaves. (b) The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections with $T$ and between neighbors.}
\end{figure}
